Machine learning topological phases in real space

### 2020-07-03 8_prb_submission/1_reply_referee/article_with_comments/article.tex
Writing section emergent information entropy wave functions

\textcolor{blue}{\citequote{eigenvector_physics} As we argue in a paragraph added to the new section \textbf{The eigenvector ensembling algorithm},  all information from a Hamiltonian can be recovered from its eigendecomposition. Therefore, by expressing a data set of eigenvectors on a suitable choice of basis (which in this paper is the real space basis), we are able to investigate properties of a family of Hamiltonians using the coordinates of eigenvectors in the chosen basis as features. Thus the use of eigenvectors greatly facilitates the exploration of model explainability techniques in data-driven physics applications while at the same time not leading to a loss of information from the systems being investigated. As to the physics being captured by our procedure, we emphasize that the Shannon entropy signatures of topological phase transitions demonstrated in our paper are physically interpretable in terms of entropic uncertainty relations which ultimately lead to the development of an uncertainty principle limiting the localizability of topological phase transitions. This topic is explored in the new section \textbf{Entropic uncertainty relations}.}

@@

\vspace{0.25cm}
\textcolor{blue}{\citequote{key_top} We agree that this sentence greatly synthesizes the results presented in the paper about the possibility of localizing topological information in SSH lattices. We have therefore moved it to the final patragraph of the section "Information entropy signatures" in the main text.}

\citequote{physical_insights} Add FFT of entropy signatures? Mention entropic uncertainty? Talk about information gain? Mention that interpretability in ML is often challenging and a fast evolving field. Furthermore, eigenvector ensembling can always be used as a preprocessing step to other ML models, such as in semi supervised learning tasks.

@@

\vspace{0.25cm}
\citequote{experiments_1_2} Experiments 1 and 2 were mentioned as subtitles to the sections.

@@

\section*{Introduction}

The quest for innovative materials that harness exotic quantum properties has lured physicists into the realm of topological insulators and topological states of matter \cite{RevModPhys.82.3045}. These materials feature previously unthought-of traits like bulk insulation coupled with metallic conductance at the surface and the splitting of currents according to spin orientation. Adding to that, these properties are protected by non-trivial topology that renders them robust to many sources of perturbation like thermal noise. Such characteristics make them promising candidates to being the cornerstone of 21st century technologies like spintronics and quantum computing.

@@

\section*{The eigenvector ensembling algorithm}\citequote{eigenvector_ensembling_alg}

\textcolor{blue}{
The eigenvector ensembling algorithm consists of five steps: 1) Generating Hamiltonians in real space and their corresponding winding numbers; 2) Creating training, validation and test sets; 3) Training on real space eigenvectors of Hamiltonians in the training set; 4) Eigenvector ensembling and 5) Bootstrapping. We describe here in detail each of these steps as they were implemented in this work.
\vspace{.3cm}
\begin{enumerate}%[topsep=0pt, partopsep=0pt]
\item[1)] \textbf{Generating Hamiltonians and winding numbers:} we start generating a number of paremeterized Hamiltonians $H(\mathbf{t})$ in real space and their corresponding winding numbers $W(\mathbf{t})$, where $\mathbf{t} = (t_1, t_2,...,t_h)$ is a vector of $h$ hopping parameters (in the simplest case of the SSH model $h$ = 2). These Hamiltonians are $N\times N$ matrices, where $N$ is twice the number of unit cells in the chain.
\item[2)] \textbf{Creating training, validation and test sets:} we split our set of parameterized Hamiltonians and winding numbers into training, validation and test sets, as is usualy done in machine learning. More explicitly, assume our hopping parameters vector $\mathbf{t}$ takes on the values $\mathbf{t}_1, \mathbf{t}_2, ..., \mathbf{t}_n$ corresponding to the Hamiltonian-winding number pairs ($H_1$, $W_1$), ..., ($H_n$, $W_n$). We partition the set \{($H_i$, $W_i$)$\mid$ $i=1,...,n$\} in three disjoint subsets: the training set, the validation set and the test set.
\item[3)] \textbf{Training on eigenvectors in real space:} since each Hamiltonian $H_i$ is represented by an $N\times N$ matrix, each one will provide $N$ eigenvectors $\mathbf{v}_i^{(1)}, \mathbf{v}_i^{(2)},...,\mathbf{v}_i^{(N)}$ to our data set. Our supervised learning algorithm of choice will take as inputs the real space eigenvectors $\mathbf{v}^{(j)}_i$ of each Hamiltonian $H_i$ in the training set and be trained to learn the winding number $W_i$ of their parent Hamiltonian $H_i$. Therefore, our dataset will consist of eigenvector-winding number pairs $(\mathbf{v}_i^{(j)}, W_i)$.
\item[4)]\textbf{Eigenvector ensembling:} in order to predict the phase of a system described by a particular Hamiltonian we need to take into account how each of its eigenvectors were classified. This amounts to performing ensemble learning on the eigenvectors of each Hamiltonian. In this work we estimate the phase probabilities for each Hamiltonian as the fraction of its eigenvectors that were classified in each phase.
\item[5)] \textbf{Bootstrapping:} We refine the phase probabilities for each Hamiltonian using a bootstrapping procedure, i.e., we repeat steps (1)-(4) $n_\text{exp}$ times, at each round sampling randomly a new training set from our grid in \textbf{t}-space. The final estimated probabilities are then arrived at by averaging the probabilities obtained in each experiment.
\end{enumerate}}

@@

\section*{Numerical experiments}

We performed two numerical experiments with the eigenvector ensembling algorithm. The first experiment deals with the simplest case, the SSH model with nearest-neighbor hopping (here called SSH 1, figure \ref{ssh1}), while the second experiment uses the SSH model with first and second nearest-neighbor hoppings (here called SSH 2, figure \ref{ssh2}).

@@

Figures \ref{figexp1_exp} and \ref{figexp2_exp} respectively illustrate single iterations of experiments 1 and 2 as seen from parameter space. The accuracy statistics presented in the following subsections and probability heatmaps shown in figures \ref{ssh1_heatmaps} and \ref{ssh2_heatmaps} were obtained after bootstrapping each experiment $n_{exp}$ = 100 times. The recovered probability heatmaps faithfully portray the phase diagrams in figure \ref{fig:phasediagrams}, with clear phase transition lines appearing in the regions of highest uncertainty.

@@

\subsection*{Experiment 1: Learning a first-neighbor hopping SSH model with decision trees}

Our test set in this experiment contained 1005 Hamiltonians (approx. 15.3\% of all data). Of the remaining 5556 Hamiltonians, 556 were randomly assigned to the training set (approx. 8.5\%) and 5000 (approx. 76.2\%) were used to compute validation scores at each iteration. These proportions between training and validation sets are such that approximately 10\%  of Hamiltonians from outside of the test set were used for training at each iteration. The composition of the train + validation set for this experiment was 50.8\% of Hamiltonians with winding number $W$ = 0 and 49.2\% with winding number $W$ = 1. The composition of the test set was 44.8\% of Hamiltonians with winding number $W$ = 0 and 55.2\% with winding number $W=1$. \textcolor{blue}{Our learning algorithm of choice for this experiment was a simple decision tree model \cite{breiman2017classification}}.\citequote{eigenvector_ensembling_alg}

@@

\begin{figure}
\centering
\subfigure[]{\label{figexp1_exp:a}\includegraphics[width=.32\textwidth]{./ssh1/periodic_100_6561/scatter_train_val_test_experiment_0.png}}\quad
\subfigure[]{\label{figexp1_exp:b}\includegraphics[width=.32\textwidth]{./ssh1/periodic_100_6561/scatter_winding_train_experiment_0.png}}
\subfigure[]{\label{figexp1_exp:c}\includegraphics[width=.32\textwidth]{./ssh1/periodic_100_6561/pcolormesh_prediction_grid_experiment_0.png}}
\caption{Visualization of a single iteration of experiment 1 as seen from 2D parameter space. (a) Train/validation/test split. (b) Distribution of winding numbers in the training set. (c) Phase diagram learned from real space lattice data by a combination of decision tree and eigenvector ensembling.}
\label{figexp1_exp}
\end{figure}

\subsection*{Experiment 2: Learning a first- and second-neighbor hoppings SSH model with random forests}
\label{exp2}

@@

\begin{figure}
\centering
\subfigure[]{\label{figexp2_exp:a}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/scatter_train_val_test_experiment_0.png}}\quad
\subfigure[]{\label{figexp2_exp:b}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/scatter_winding_train_experiment_0.png}}
\subfigure[]{\label{figexp2_exp:c}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/pcolormesh_prediction_grid_experiment_0.png}}
\caption{Visualization of a single iteration of experiment 2 as seen from 2D parameter space. (a) Train/validation/test split. (b) Distribution of winding numbers in the training set. (c) Phase diagram learned from real space lattice data by a combination of random forest and eigenvector ensembling.}
\label{figexp2_exp}
\end{figure}

\section*{Information entropy signatures}

@@

The (normalized) relevance of a feature is given by how much it reduces a loss function (e.g. Shannon information entropy or Gini impurity \cite{friedman2001elements}). By averaging normalized relevances as measured by reduction in the information entropy of ensembles of real space eigenvectors across $n_{exp}$ = 100 iterations of both experiment 1 and experiment 2 we recovered Shannon entropy signals that reveal which lattice sites were consistently more relevant in learning topological phases from data in real space for each experiment. These signals are the information entropy signatures of each topological phase transition.

The bar plots in figure \ref{feature_importances} show how informative each lattice site was in learning topological phases for experiments 1 and 2. They represent the information entropy signatures along the lattices in each SSH system. For experiment 1, only six lattice sites $S_1$ = (0, 1, 3, 50, 51, 53) corresponding to the two sharp peaks seen in figure \ref{feature_importances_ssh1} contributed approximately 70\% of total reduction in Shannon entropy. Similarly, approximately 30\% of total reduction in the Shannon information entropy of eigenvector data from experiment 2 was achieved by eighteen lattice sites $S_2$ = (0, 1, 2, 3, 4, 5, 46, 48, 49, 50, 51, 53, 94, 95, 96, 97, 98, 99) distributed along the three peaks in figure \ref{feature_importances_ssh2}.

Each of the information entropy signatures shown in figure \ref{feature_importances} captures a general pattern that persists regardless of the length of the lattice (i.e, number of unit cells) used to compute them. They are not, therefore, artifacts of particular choices of hyperparameters used to run the eigenvector ensembling algorithm. We present the information entropy signatures for longer lattices in the section \textbf{Information entropy signatures in the macroscopic limit} in the Supplementary Material.

To see if learning the topological phases can be achieved efficiently by employing simpler models we reran experiments 1 and 2 using only a small subset of most relevant lattice sites. In our rerun of experiment 1 using only lattice sites $S'_1$ = (0, 50, 51, 99) (which contributed approximately 45 \% of total reduction in Shannon information entropy in experiment 1), mean accuracies were 97.0\% for training eigenvectors, 91.5\% for validation eigenvectors and 72.8\% for test eigenvectors. Mean accuracies obtained from eigenvector ensembling were 99.1\% for training Hamiltonians, 99.5\% for validation Hamiltonians and 94.5\% for test Hamiltonians.

Mean accuracies for our rerun of experiment 2 using only lattice sites $S'_2$ = (0, 1, 3, 48, 50, 51, 96, 98, 99) (which contributed approximately 20 \% of total reduction in Shannon information entropy in experiment 2) were 99.9\% for training eigenvectors, 87.7\% for validation eigenvectors and 47.3\% for test eigenvectors. Eigenvector ensembling yields mean accuracies of 100\% for training Hamiltonians, 99.5\% for validation Hamiltonians and 74.5\% for test Hamiltonians.

@@

\begin{figure}
\centering
\subfigure[]{\label{ssh1_heatmap_0}\includegraphics[width=.32\textwidth]{./ssh1/periodic_100_6561/imshow_winding_grid_winding_0_sim.png}}
\subfigure[]{\label{ssh1_heatmap_1}\includegraphics[width=.32\textwidth]{./ssh1/periodic_100_6561/imshow_winding_grid_winding_1_sim.png}}
\subfigure[]{\label{ssh1_heatmap}\includegraphics[width=.32\textwidth]{./ssh1/periodic_100_6561/merge_imshow_winding_grids_second_sim.png}}
\caption{Probability heatmaps learned by a combination of decision trees with eigenvector ensembling from bulk real space eigenvector data in experiment 1. Heatmaps were averaged across all 100 iterations of the experiment. (a) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 0.  (b) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 1. (c) The phase diagram resulting from heatmaps (a) and (b).}
\label{ssh1_heatmaps}
\end{figure}

\begin{figure}
\centering
\subfigure[]{\label{ssh2_heatmap_-1}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/imshow_winding_grid_winding_-1_sim.png}}
\subfigure[]{\label{ssh2_heatmap_0}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/imshow_winding_grid_winding_0_sim.png}}
\subfigure[]{\label{ssh2_heatmap_1}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/imshow_winding_grid_winding_1_sim.png}}
\subfigure[]{\label{ssh2_heatmap_2}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/imshow_winding_grid_winding_2_sim.png}}
\subfigure[]{\label{ssh2_heatmap}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/merge_imshow_winding_grids_sim.png}}
\caption{Probability heatmaps learned by a combination of random forests with eigenvector ensembling from bulk real space eigenvector data in experiment 2. Heatmaps were averaged across all 100 iterations of the experiment. (a) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to -1.  (b) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 0. (c) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 1. (d) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 2. (e) The phase diagram resulting from heatmaps (a)-(d).}
\label{ssh2_heatmaps}
\end{figure}

\section*{Discussion}

Given the increasing complexity of systems studied in condensed matter physics and the rising demand for materials with exotic and robust properties to power future technological progress, it is only expected that data-driven approaches to physics will grow in demand. Our work represents a step in this direction, as we have implemented a data-driven approach to the search for new topological materials bypassing the use of wavevector space data.

>>>>

\textcolor{blue}{\citequote{eigenvector_physics} As we argue in a paragraph added to the new section \textbf{The eigenvector ensembling algorithm},  all information from a Hamiltonian can be recovered from its eigendecomposition. Therefore, by expressing a data set of eigenvectors on a suitable choice of basis (which in this paper is the real space basis), we are able to investigate properties of a family of Hamiltonians using the coordinates of eigenvectors in the chosen basis as features. Thus the use of eigenvectors greatly facilitates the exploration of model explainability techniques in data-driven physics applications while at the same time not leading to a loss of information from the systems being investigated. As to the physics being captured by our procedure, see the answer to the issue \citequote{physical_insights}.}

@@

\vspace{0.25cm}
\textcolor{blue}{\citequote{key_top} We agree that this sentence greatly synthesizes the results presented in the paper about the possibility of localizing topological information in SSH lattices. We have therefore moved it to the final patragraph of the section \textbf{Information entropy signatures} in the main text.}

\citequote{physical_insights} We have added the new section \textbf{Emergent information entropy wave functions} in which we interpret an information entropy signature as a probability distribution arising from an emergent information entropy wave function in the continuum limit of an infinit lattice. Therefore, the paper now performs a full cycle of data-driven physics, as we propose a method to investigate topological phase transitions using real space eigenvector data from simulations of SSH systems (the eigenvector ensembling algorithm), use model explainability to extract new, previously unknown properties of the SSH systems (the information entropy signatures) and finally interpret these new properties in the light of theoretical physics. It is important to point that the majority of physics papers published to date on the use of machine learning in the investigation of condensed matter systems have been proofs of concept aimed at showing that the machine learning algorithms that have become ubiquitously famous during the last decade indeed are able to recognize the known relevant patterns associated with specific physical phenomena. Our paper goes beyond this task as we use machine learning and model explainability to discover new properties of the physical systems under consideration and (with the addition of this new section) map these previously unknown properties into new theoretical  constructs.

@@

\vspace{0.25cm}
\citequote{experiments_1_2} We have added references to the relevant subsections whenever each numerical experiment is mentioned outside of its subsection throughout the text. The designation "experiment 1" is used to abbreviate the subsection \textbf{Experiment 1: Learning a first-neighbor hopping SSH model with decision trees}. Similarly, the designation "experiment 2" is used to abbreviate the subsection \textbf{Experiment 2: Learning a first- and second-neighbor hoppings SSH model with random forests}.

@@

\section{Introduction}
\label{introduction}
The quest for innovative materials that harness exotic quantum properties has lured physicists into the realm of topological insulators and topological states of matter \cite{RevModPhys.82.3045}. These materials feature previously unthought-of traits like bulk insulation coupled with metallic conductance at the surface and the splitting of currents according to spin orientation. Adding to that, these properties are protected by non-trivial topology that renders them robust to many sources of perturbation like thermal noise. Such characteristics make them promising candidates to being the cornerstone of 21st century technologies like spintronics and quantum computing.

@@

\section{The eigenvector ensembling algorithm}\citequote{eigenvector_ensembling_alg}
\label{the_eigenvector_ensembling_algorithm}
\textcolor{blue}{
The eigenvector ensembling algorithm consists of five steps: 1) Generating Hamiltonians in real space and their corresponding winding numbers; 2) Creating training, validation and test sets; 3) Training on real space eigenvectors of Hamiltonians in the training set; 4) Eigenvector ensembling and 5) Bootstrapping. We describe here in detail each of these steps as they were implemented in this work.
\vspace{.3cm}
\begin{enumerate}%[topsep=0pt, partopsep=0pt]
\item[1)] \textbf{Generating Hamiltonians and winding numbers:} we start generating a number of paremeterized Hamiltonians $H(\mathbf{t})$ in real space and their corresponding winding numbers $W(\mathbf{t})$, where $\mathbf{t} = (t_1, t_2,...,t_h)$ is a vector of $h$ hopping parameters (in the simplest case of the SSH model $h$ = 2). These Hamiltonians are $N\times N$ matrices, where $N$ is twice the number of unit cells in the chain.
\item[2)] \textbf{Creating training, validation and test sets:} we split our set of parameterized Hamiltonians and winding numbers into training, validation and test sets, as is usualy done in machine learning. More explicitly, assume our hopping parameters vector $\mathbf{t}$ takes on the values $\mathbf{t}_1, \mathbf{t}_2, ..., \mathbf{t}_n$ corresponding to the Hamiltonian-winding number pairs ($H_1$, $W_1$), ..., ($H_n$, $W_n$). We partition the set \{($H_i$, $W_i$)$\mid$ $i=1,...,n$\} in three disjoint subsets: the training set, the validation set and the test set.
\item[3)] \textbf{Training on eigenvectors in real space:} since each Hamiltonian $H_i$ is represented by an $N\times N$ matrix, each one will provide $N$ eigenvectors $\mathbf{v}_i^{(1)}, \mathbf{v}_i^{(2)},...,\mathbf{v}_i^{(N)}$ to our data set. Our supervised learning algorithm of choice will take as inputs the real space eigenvectors $\mathbf{v}^{(j)}_i$ of each Hamiltonian $H_i$ in the training set and be trained to learn the winding number $W_i$ of their parent Hamiltonian $H_i$. Therefore, our dataset will consist of eigenvector-winding number pairs $(\mathbf{v}_i^{(j)}, W_i)$.
\item[4)]\textbf{Eigenvector ensembling:} in order to predict the phase of a system described by a particular Hamiltonian we need to take into account how each of its eigenvectors were classified. This amounts to performing ensemble learning on the eigenvectors of each Hamiltonian. In this work we estimate the phase probabilities for each Hamiltonian as the fraction of its eigenvectors that were classified in each phase.
\item[5)] \textbf{Bootstrapping:} We refine the phase probabilities for each Hamiltonian using a bootstrapping procedure, i.e., we repeat steps (1)-(4) $n_\text{exp}$ times, at each round sampling randomly a new training set from our grid in \textbf{t}-space. The final estimated probabilities are then arrived at by averaging the probabilities obtained in each experiment.
\end{enumerate}}

@@

\section{Numerical experiments}
\label{numerical_experiments}
We performed two numerical experiments with the eigenvector ensembling algorithm. The first experiment deals with the simplest case, the SSH model with nearest-neighbor hopping (here called SSH 1, figure \ref{ssh1}), while the second experiment uses the SSH model with first and second nearest-neighbor hoppings (here called SSH 2, figure \ref{ssh2}).

@@

Figures \ref{figexp1_exp} and \ref{figexp2_exp} respectively illustrate single iterations of experiment 1 (section \ref{experiment_1}) and experiment 2 (section \ref{experiment_2}) as seen from parameter space. The accuracy statistics presented in the following subsections and probability heatmaps shown in figures \ref{ssh1_heatmaps} and \ref{ssh2_heatmaps} were obtained after bootstrapping each experiment $n_{exp}$ = 100 times. The recovered probability heatmaps faithfully portray the phase diagrams in figure \ref{fig:phasediagrams}, with clear phase transition lines appearing in the regions of highest uncertainty.

@@

\subsection{Experiment 1: Learning a first-neighbor hopping SSH model with decision trees}
\label{experiment_1}
Our test set in this experiment contained 1005 Hamiltonians (approx. 15.3\% of all data). Of the remaining 5556 Hamiltonians, 556 were randomly assigned to the training set (approx. 8.5\%) and 5000 (approx. 76.2\%) were used to compute validation scores at each iteration. These proportions between training and validation sets are such that approximately 10\%  of Hamiltonians from outside of the test set were used for training at each iteration. The composition of the train + validation set for this experiment was 50.8\% of Hamiltonians with winding number $W$ = 0 and 49.2\% with winding number $W$ = 1. The composition of the test set was 44.8\% of Hamiltonians with winding number $W$ = 0 and 55.2\% with winding number $W=1$. \textcolor{blue}{Our learning algorithm of choice for this experiment was a simple decision tree model \cite{breiman2017classification}}.\citequote{eigenvector_ensembling_alg}

@@

\begin{figure}
\centering
\subfigure[]{\label{figexp1_exp:a}\includegraphics[width=.32\textwidth]{./ssh1/periodic_100_6561/scatter_train_val_test_experiment_0.png}}\quad
\subfigure[]{\label{figexp1_exp:b}\includegraphics[width=.32\textwidth]{./ssh1/periodic_100_6561/scatter_winding_train_experiment_0.png}}
\subfigure[]{\label{figexp1_exp:c}\includegraphics[width=.32\textwidth]{./ssh1/periodic_100_6561/pcolormesh_prediction_grid_experiment_0.png}}
\caption{Visualization of a single iteration of experiment 1 (section \ref{experiment_1}) as seen from 2D parameter space. (a) Train/validation/test split. (b) Distribution of winding numbers in the training set. (c) Phase diagram learned from real space lattice data by a combination of decision tree and eigenvector ensembling.}
\label{figexp1_exp}
\end{figure}

\subsection{Experiment 2: Learning a first- and second-neighbor hoppings SSH model with random forests}
\label{experiment_2}

@@

\begin{figure}
\centering
\subfigure[]{\label{figexp2_exp:a}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/scatter_train_val_test_experiment_0.png}}\quad
\subfigure[]{\label{figexp2_exp:b}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/scatter_winding_train_experiment_0.png}}
\subfigure[]{\label{figexp2_exp:c}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/pcolormesh_prediction_grid_experiment_0.png}}
\caption{Visualization of a single iteration of experiment 2 (section \ref{experiment_2}) as seen from 2D parameter space. (a) Train/validation/test split. (b) Distribution of winding numbers in the training set. (c) Phase diagram learned from real space lattice data by a combination of random forest and eigenvector ensembling.}
\label{figexp2_exp}
\end{figure}

\section{Information entropy signatures}
\label{information_entropy_signatures}

@@

The (normalized) relevance of a feature is given by how much it reduces a loss function (e.g. Shannon information entropy or Gini impurity \cite{friedman2001elements}). By averaging normalized relevances as measured by reduction in the information entropy of ensembles of real space eigenvectors across $n_{exp}$ = 100 iterations of both experiment 1 (section \ref{experiment_1}) and experiment 2 (section \ref{experiment_2}) we recovered Shannon entropy signals that reveal which lattice sites were consistently more relevant in learning topological phases from data in real space for each experiment. These signals are the information entropy signatures of each topological phase transition.

The bar plots in figure \ref{feature_importances} show how informative each lattice site was in learning topological phases for each experiment. They represent the information entropy signatures along the lattices in each SSH system. For experiment 1 (section \ref{experiment_1}), only six lattice sites $S_1$ = (0, 1, 3, 50, 51, 53) corresponding to the two sharp peaks seen in figure \ref{feature_importances_ssh1} contributed approximately 70\% of total reduction in Shannon entropy. Similarly, approximately 30\% of total reduction in the Shannon information entropy of eigenvector data from experiment 2 (section \ref{experiment_2}) was achieved by eighteen lattice sites $S_2$ = (0, 1, 2, 3, 4, 5, 46, 48, 49, 50, 51, 53, 94, 95, 96, 97, 98, 99) distributed along the three peaks in figure \ref{feature_importances_ssh2}.

\textcolor{blue}{We should remark on the subtleties of the information entropy signatures presented here. Although they give us a visualization of how important each lattice site was in determining the topological phases of Hamiltonians, they actually express a global property of the whole lattice. In the next section, where we develop a quantum formalism for the information entropy signatures discovered in this section, these seemingly antagonistic conceptions shall be harmonized. What is important to emphasize at this point is that the information entropy signature should not be naively taken at face value: a lattice site that appears unimportant in an information entropy signature plot may not be unimportant or void of topological information by itself.}

\textcolor{blue}{To give a concrete example, reduction in Shannon entropy tends to be distributed among highly correlated variables. This implies that if only a single lattice site in a highly correlated subset is used by a learning algorithm, it will likely inherit most of the reduction in Shannon entropy from the other correlated lattice sites that were not taken into account by the algorithm. The corollary of this fact is that lattice sites that carry redundant information that is also available from other lattice sites tend to have decreased importance in the information entropy signature.  In this regard the information entropy signatures presented here express a summary of relations between lattice sites and are therefore intrinsically global.}

%Another important point to be emphasized is that these signals were recovered from the analysis of data from several thousand SSH systems in %each experiment and therefore they are not a property of a single SSH lattice. They are rather a pattern that emerges from correlating %topological winding numbers with lattice eigenvector data for several SSH systems.

\textcolor{blue}{Each of the information entropy signatures shown in figure \ref{feature_importances} captures a general pattern that persists regardless of the length of the lattice (i.e., the number of unit cells) used to compute them. In fact, by rerunning each experiment with longer lattices we have verified that the signals in figures \ref{feature_importances_ssh1} and \ref{feature_importances_ssh2} appear to converge to well defined continuous density functions in the macroscopic limit. They are not, therefore, artifacts of particular choices of hyperparameters used to run the eigenvector ensembling algorithm. The information entropy signatures for longer lattices are presented in the section \textbf{Numerical explorations on longer lattices} in the Supplementary Material.}

%The macroscopic limit of these information entropy signals may be an important signature of topological systems and thus merits further %theoretical investigation. This is pursued in the next section. We present the information entropy signatures for longer lattices in the %section \textbf{Numerical explorations on longer lattices} in the Supplementary Material.}

To see if learning the topological phases can be achieved efficiently by employing simpler models we reran each experiment using only a small subset of most relevant lattice sites. In our rerun of experiment 1 (section \ref{experiment_1}) using only lattice sites $S'_1$ = (0, 50, 51, 99) (which contributed approximately 45 \% of total reduction in Shannon information entropy), mean accuracies were 97.0\% for training eigenvectors, 91.5\% for validation eigenvectors and 72.8\% for test eigenvectors. Mean accuracies obtained from eigenvector ensembling were 99.1\% for training Hamiltonians, 99.5\% for validation Hamiltonians and 94.5\% for test Hamiltonians.

Mean accuracies for our rerun of experiment 2 (section \ref{experiment_2}) using only lattice sites $S'_2$ = (0, 1, 3, 48, 50, 51, 96, 98, 99) (which contributed approximately 20 \% of total reduction in Shannon information entropy) were 99.9\% for training eigenvectors, 87.7\% for validation eigenvectors and 47.3\% for test eigenvectors. Eigenvector ensembling yields mean accuracies of 100\% for training Hamiltonians, 99.5\% for validation Hamiltonians and 74.5\% for test Hamiltonians.

@@

\begin{figure}
\centering
\subfigure[]{\label{ssh1_heatmap_0}\includegraphics[width=.32\textwidth]{./ssh1/periodic_100_6561/imshow_winding_grid_winding_0_sim.png}}
\subfigure[]{\label{ssh1_heatmap_1}\includegraphics[width=.32\textwidth]{./ssh1/periodic_100_6561/imshow_winding_grid_winding_1_sim.png}}
\subfigure[]{\label{ssh1_heatmap}\includegraphics[width=.32\textwidth]{./ssh1/periodic_100_6561/merge_imshow_winding_grids_second_sim.png}}
\caption{Probability heatmaps learned by a combination of decision trees with eigenvector ensembling from bulk real space eigenvector data in experiment 1 (section \ref{experiment_1}). Heatmaps were averaged across all 100 iterations of the experiment. (a) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 0.  (b) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 1. (c) The phase diagram resulting from heatmaps (a) and (b).}
\label{ssh1_heatmaps}
\end{figure}

\begin{figure}
\centering
\subfigure[]{\label{ssh2_heatmap_-1}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/imshow_winding_grid_winding_-1_sim.png}}
\subfigure[]{\label{ssh2_heatmap_0}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/imshow_winding_grid_winding_0_sim.png}}
\subfigure[]{\label{ssh2_heatmap_1}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/imshow_winding_grid_winding_1_sim.png}}
\subfigure[]{\label{ssh2_heatmap_2}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/imshow_winding_grid_winding_2_sim.png}}
\subfigure[]{\label{ssh2_heatmap}\includegraphics[width=.32\textwidth]{./ssh2/periodic_100_6561/merge_imshow_winding_grids_sim.png}}
\caption{Probability heatmaps learned by a combination of random forests with eigenvector ensembling from bulk real space eigenvector data in experiment 2 (section \ref{experiment_2}). Heatmaps were averaged across all 100 iterations of the experiment. (a) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to -1.  (b) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 0. (c) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 1. (d) Probability heatmap showing the probability that a Hamiltonian in the grid has winding number equal to 2. (e) The phase diagram resulting from heatmaps (a)-(d).}
\label{ssh2_heatmaps}
\end{figure}

\textcolor{blue}{
\section{Emergent information entropy wave functions}
\label{emergent_information_entropy_wave_functions}
}
\textcolor{blue}{
The information entropy signatures that we have seen in the previous section pose an immediate theoretical angst: how can a signal that is locally defined (information entropy signature) arise from a global property (topological invariant) of the whole SSH lattice?    
In this final section we venture into a theoretical exploration of the information entropy signatures in the hope of elucidating this issue. Our goal is to provide a theoretical framework to interpret the new concepts brought to light in the previous section.  The discussion presented here will be complemented by the numerical experiments in the section \textbf{Numerical explorations on longer lattices} in the Supplementary Material.}

\textcolor{blue}{
We can think of the Shannon information entropy signatures in figures \ref{feature_importances_ssh1} and \ref{feature_importances_ssh2} as discrete information entropy mass functions that, in the continuum (i.e. macroscopic) limit of an infinite chain, lead to local entropy density functions along the lattices. By mapping the lattice to a Riemann partition of the closed interval $[0,1]$, the cumulative distribution of topological information along a lattice will be given by}

\begin{equation}
\label{entropy_density}
F_S(x) = \int_{0}^x\rho_S(x')dx'
\end{equation}
where $x \in [0,1]$ is a spatial coordinate along the lattice and $\rho_S(x)$ is the local Shannon information entropy density function in the continuum limit.

\textcolor{blue}{
Given the quantum nature of the phase transitions being discussed, the information entropy density function $\rho_S(x)$ can be naturally interpreted as the squared magnitude of a spatial Shannon information entropy wave function,
}
\begin{equation}
\label{shannon_wave_function}
\rho_S(x) = |\psi_S(x)|^2,
\end{equation}
the amount of topological information available from a single point in the continuous 1D string then being expressed in bra-ket notation by Born's rule,

\begin{equation}
\label{born_rule_spatial}
\rho_S(x) = |\braket{x|\psi_S}|^2.
\end{equation}

\textcolor{blue}{
The interpretation of the information entropy signatures of the previous section in terms of information entropy wave functions opens several avenues for theoretical investigation.  Here we explore its most forthright corollary, which is the establishment of uncertainty relations for topological phase transitions. 
}

\textcolor{blue}{
The counterpart of the spatial information entropy wave function $\psi_S(x)$ in wavevector space is its Fourier transform }
\begin{equation}
\label{fourier_shannon_wave_function}
\hat{\psi}_S(k) = \int_{0}^{1}\psi_S(x)e^{-2\pi ikx}dx
\end{equation}
\textcolor{blue}{
from which the Shannon information entropy density function can be computed in wavevector space,}
\begin{equation}
\label{shannon_wave_function_wavevector_space}
\hat{\rho}_S(k) = |\hat{\psi}_S(k)|^2.
\end{equation}
\textcolor{blue}{
Let us further denote the mean, the variance and the entropy of $\rho_S$ by $\bar{\rho}_S$, $\Delta \rho_S^2 $ and $H[\rho_S]$ respectively. Explicitly, we have }
\begin{subequations}
\label{statistics}
\begin{equation}
\label{mean}
\bar{\rho}_S = \int_0^1x\rho_S(x)dx,
\end{equation}
\begin{equation}
\label{variance}
\Delta\rho_S^2 = \int_0^1(x-\bar{\rho}_S)^2\rho_S(x)dx,\quad \text{and}
\end{equation}
\begin{equation}
\label{entropy}
H[\rho_S] = \int_0^1\rho_S(x)\ln\big(\rho_S(x)\big) dx,
\end{equation}
\end{subequations}
\textcolor{blue}{
with analogous equations for the wavevector space counterparts $\bar{\hat{\rho}}_S$, $\Delta \hat{\rho}_S^2 $ and $H[\hat{\rho_S}]$ in terms of $k$ and $\hat{\rho}$.}

\textcolor{blue}{
In possession of this quantum formalism, we may write topological versions of two canonical uncertainty relations that bind together the real space and wavevector space information entropy density functions \eqref{shannon_wave_function} and \eqref{shannon_wave_function_wavevector_space}: }
